---
title: "Pre processing"
author: "Yanchao"
date: '2019-03-25'
output: github_document
---
## Load the library
```{r,warning=FALSE}
library(RColorBrewer)
library(cluster)
library(pvclust)
library(xtable)
library(limma)
library(plyr)
library(lattice)
library(RCurl)
options(download.file.method = "curl")
library(GEOquery)
library(knitr)
library(pheatmap)
  library(stringr)
library(ggplot2)
library(dplyr)
library(reshape2)
library(tidyverse)
library(mice)
library(VIM)
```

## Load the data

### geo_GSE18123 data
```{r,warning=FALSE}
 geo_GSE18123 <- getGEO("GSE18123", GSEMatrix = TRUE)
geo_GSE18123<- geo_GSE18123[[1]]
```

### Get expression data of geo_GSE18123
```{r}
#Get expression data  
data_GSE18123<-exprs(geo_GSE18123)
hist(data_GSE18123, col = "gray", main = "GSE70213 - Histogram")

```

It appears a lot of genes have values << 500000. 
We consider taking Log2 transformation.

```{r}
hist(log2(data_GSE18123 + 1), col = "gray", main = "GSE70213 log transformed - Histogram")
log_data_GSE18123<-log2(data_GSE18123 + 1)
log_data_GSE18123<-as.data.frame(log_data_GSE18123)
```

### get melta data of GSE18123
```{r}
## get melta data of GSE18123
prDes_GSE18123 <- pData(geo_GSE18123)[,c("organism_ch1","title",colnames(pData(geo_GSE18123))[grep("characteristics", colnames(pData(geo_GSE18123)))])]
meta_data_GSE18123<-prDes_GSE18123[,1:5]
colnames(meta_data_GSE18123) = c("organism","sample_name","diagnosis","gender","age")
meta_data_GSE18123$diagnosis = as.factor(gsub("diagnosis: ","", meta_data_GSE18123$diagnosis))

meta_data_GSE18123$age = gsub("age: ","", meta_data_GSE18123$age)

meta_data_GSE18123$age<-as.integer(str_extract(meta_data_GSE18123$age, "[0-9]{2,3}"))
meta_data_GSE18123$diagnosis
meta_data_GSE18123$age <- meta_data_GSE18123$age/12

meta_data_GSE18123$diagnosis<-ifelse(meta_data_GSE18123$diagnosis == "PDD-NOS", "AUTISM", ifelse(meta_data_GSE18123$diagnosis == "ASPERGER'S DISORDER", "AUTISM",  ifelse(meta_data_GSE18123$diagnosis == "CONTROL", "CONTROL", ifelse(meta_data_GSE18123$diagnosis == "AUTISM", "AUTISM", "error"))))
meta_data_GSE18123$batch<-"none"

kable(head(meta_data_GSE18123))

```

```{r}
dim(meta_data_GSE18123)
```

### geo_GSE25507 data
```{r,warning=FALSE}
# 
geo_GSE25507 <- getGEO("GSE25507", GSEMatrix = TRUE)
geo_GSE25507<- geo_GSE25507[[1]]
```

### Get expression data of GSE25507
```{r}
#Get expression data of GSE25507 
data_GSE25507<-exprs(geo_GSE25507)
hist(data_GSE25507, col = "gray", main = "GSE25507 - Histogram")
```

It appears a lot of genes have values < 1000. 
```{r}
hist(log2(data_GSE25507 + 1), col = "gray", main = "GSE25507 log transformed - Histogram")
log_data_GSE25507<-log2(data_GSE25507 + 1)
log_data_GSE25507<-as.data.frame(log_data_GSE25507)
```

### get meta data of GSE25507
```{r}
# get meta data of GSE25507
prDes_GSE25507 <- pData(geo_GSE25507)[,c("organism_ch1","title",colnames(pData(geo_GSE18123))[grep("characteristics", colnames(pData(geo_GSE25507)))])]
meta_data_GSE25507<-prDes_GSE25507[,1:5]
colnames(meta_data_GSE25507) = c("organism","sample_name","batch","diagnosis","age")
meta_data_GSE25507$diagnosis = as.factor(gsub("diagnosis: ","", meta_data_GSE25507$diagnosis))

meta_data_GSE25507$age = gsub("age: ","", meta_data_GSE25507$age)

meta_data_GSE25507$age<-as.integer(str_extract(meta_data_GSE25507$age, "[0-9]{1}"))
meta_data_GSE25507$diagnosis<-ifelse(meta_data_GSE25507$diagnosis == "group: control", "CONTROL", ifelse(meta_data_GSE25507$diagnosis == "group: autism", "AUTISM", "error"))
meta_data_GSE25507$batch<-ifelse(meta_data_GSE25507$batch == "scan batch: Batch 1", "batch 1", ifelse(meta_data_GSE25507$batch == "scan batch: Batch 2", "batch 2", "error"))
kable(head(meta_data_GSE25507))
```

```{r}
dim(meta_data_GSE25507)

```

## Imputing the missing data by multiple imputation
```{r}
## identify how many NA in the data
sum(is.na(meta_data_GSE25507$age))
```

```{r}
set.seed(2019)
imputed_Data <- mice(meta_data_GSE25507, m=5, maxit = 50, method = 'pmm', seed = 500)
## We use the third iteration:
meta_data_GSE25507_2 <- complete(imputed_Data, 3)
rownames(meta_data_GSE25507_2) <- rownames(meta_data_GSE25507)
```


### Combine two meta data
```{r}
F_meta_data_GSE18123<-meta_data_GSE18123 %>% dplyr::select(organism,sample_name,diagnosis,age, batch)
F_meta_data_GSE25507<-meta_data_GSE25507_2 %>% dplyr::select(organism,sample_name,diagnosis,age, batch)
## Combine two meta data
Meta_data = rbind(F_meta_data_GSE18123, F_meta_data_GSE25507)
```

## density plot
```{r}
# density plot

dat.geneMeans <- c(rowMeans(log_data_GSE25507), rowMeans(log_data_GSE18123)) 
plotDat <- data.frame(mean_gene = dat.geneMeans,
                      Dataset = rep(c('log_data_GSE25507', 'log_data_GSE18123'), each = nrow(log_data_GSE25507)))

(probeAvg <- ggplot(data = plotDat, aes(x = mean_gene, col = Dataset)) +
   geom_density() + 
   ggtitle("Average gene expression value density of two experiments") + 
   xlab("mean of gene ") + 
   ylab("Density") + 
   theme_bw()
)


```

## Quantile normalization
```{r}
# combine data from two experiments into one matrix, each column represents gene expression values of one sample
combine_matrix <- as.matrix(cbind(log_data_GSE18123,log_data_GSE25507))
str(combine_matrix, max.level = 0)
```
```{r}
# quantile normalization
system.time(combine_norm <- normalizeBetweenArrays(combine_matrix))
```
```{r}
dat.geneMeans <- c(rowMeans(combine_norm[, 1:ncol(log_data_GSE18123)]), rowMeans(combine_norm[, ncol(log_data_GSE18123):ncol(combine_norm)])) 
plotDat2 <- data.frame(mean_gene = dat.geneMeans,
                      Dataset = rep(c('log_data_GSE25507', 'log_data_GSE18123'), each = nrow(log_data_GSE25507)))

(probeAvg <- ggplot(data = plotDat2, aes(x = mean_gene, col = Dataset)) +
   geom_density() + 
   ggtitle("Average gene expression value density of two experiments") + 
   xlab("mean of gene ") + 
   ylab("Density") + 
   theme_bw()
)

```

## Save the data to avoid future re-downloading
```{r}
Meta_data$diagnosis<- Meta_data$diagnosis %>% factor(levels = c("AUTISM", "CONTROL"))
Meta_data$batch<- Meta_data$batch %>% factor(levels = c("none", "batch 1","batch 2"))
#Saving normalized data seperately
saveRDS(combine_norm, file = "combine_norm.rds")
saveRDS(Meta_data, file = "Meta_data.rds")
```
